Exploring the Relationship between Family Background and Children's Education Level in the United States (2021)

### 2023-04-16 outputs/paper/paper.qmd

---
title: "Exploring the Relation between Family Background and Children's Education Level in the United States (2021)"
author: 
  - Ruibo Sun
thanks: "Code and Data: https://github.com/ruibosun/revised-how-parents-affect-childrens-education"
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
abstract: "This paper explores the relationship between family status variables, such as parents' socio-economic status and educational level, and their impact on children's academic achievement and education level using the data from 2021 General Social Survey (GSS). Previous studies have suggested that parents' education can positively influence their children's academic performance due to increased access to resources and greater involvement in their education. However, recent research has demonstrated that this relationship is complex and can be influenced by a variety of factors. The findings in this paper suggest that parents' education, family income, and prestige score all affect the children's education level. Further research is needed to fully understand the complexities of this relationship and use the results to improve educational policies and programs."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
library(broom)
library(knitr)
library(kableExtra)
library(tidyverse)
library(dplyr)
library(here)
library(haven)
library(scales)
#library(purrr)
#library(forcats)
library(patchwork)
```

@@

In short, the findings in this paper demonstrate that parents’ education has a positive correlation with children’s education. Additionally, the region of living during the teenage years can also impact children’s education. Parents’ prestige score seems to be another factor in children’s education. However, there is some limitation on the definition of prestige score that may influence the overall results. This will be discussed in the @sec-weakness.

The data will be presented clearly and succinctly with plots and tables. This following analysis is processed in R [@R]  with packages of tidyverse [@tidyverse], dplyr [@dplyr], here [@here], haven [@haven] and broom [@broom]. The tables are constructed via knitr [@knitr], scales [@scales] and kableExtra [@kableExtra]. The package inside tidyverse helps to create the plots in ggplot2 [@tidyverse]. The missing values were imputed using the mice package [@mice]. This paper is knitted as a PDF file by the packages of R markdown [@RMarkdown] and formatted using patchwork [@patchwork].

@@

The target population is US adults aged 18 and older. The GSS uses a multistage probability sampling approach to recruit its sample from the US Census Bureau's Master Address File. One strength of the GSS is its long history and large sample size, but a potential weakness is a reliance on self-reported data, which can be subject to response bias. The questionnaire includes both closed-ended and open-ended questions, but there may be concerns about biased or leading questions. The GSS employs various methods to adjust for non-response and ensure the representativeness of the sample. The questionnaire includes both closed-ended and open-ended questions, but there may be concerns about biased or leading questions. [@norcuofch] The GSS employs various methods to adjust for non-response and ensure the representativeness of the sample. The limitation will be explained in detail in the @sec-weakness.

$$
Prestige_{ij}=\mu+\alpha_i+\beta_{j}+\epsilon_{ij}
$$ {#eq-modpr}

This paper will test and explore how parents’ social-economics status can affect children’s education. The data that will be used in this paper comes from the US General Social Survey from the National Opinion Research Center at the University of Chicago. In this study, the "children" are the respondents and "parent" are the parents of the respondents.

The factor of socio-economics status is measured using the occupation prestige score. It is a measure used in social science research to assess the level of social status or prestige associated with a particular occupation or profession, which was developed by the National Opinion Research Center at the University of Chicago. As part of the GSS, respondents are asked to rate the prestige or social standing of various occupations on a scale of 1 to 100, with higher scores indicating greater prestige. The ideal dataset would have 90,090 ratings from 1,001 raters, but due to various issues, only 82,800 ratings from 946 raters were used. Invalid or inconsistent ratings, such as cases with reversed codes, too few ratings, equal scores for all occupations, or a low standard deviation of ratings, were excluded to ensure the reliability of the occupational scores.[@smith]

Next,the prestige score is obtained by the @eq-modpr. $Prestige_{ij}$ is the prestige score for occupational title i and rater j. The $\alpha_i$ represents the occupational differences of interest, while the $\beta_j$ represents differences among the raters that need to be controlled for in estimating the $\alpha_i$. The $\mu$ is the overall mean prestige score, and $\epsilon_{ij}$ represents the random error term. The range of the prestige scores is from 0 to 100, with higher scores indicating a better or more prestigious occupation[@smith].

@@

- **mapres10**: A numeric variable indicating the respondent's mother’s prestige score.
- **papres10**: A numeric variable indicating the respondent's father’s prestige score.
- **paeduc**: A numeric variable indicating the number of years of education completed by the respondent's father.
- **maeduc**: A numeric variable indicating the number of years of education completed by the respondent's mother.
- **educ**: A numeric variable indicating the number of years of education completed by the respondent.
- **degree**: A categorical variable indicating the highest degree earned by the respondent.
- **born**: A categorical variable indicating the region of the US where the respondent was born.
- **reg16**: A categorical variable indicating the region of the US where the respondent lived at age of 16.
- **incom16**: A categorical variable indicating the respondent's total household at age of 16.

@@

As mentioned before, this paper is an revised version of [the previous paper](https://github.com/ruibosun/how-parents-affect-childrens-education/blob/main/outputs/paper/paper.pdf) which has some limitations and weaknesses. In order to improve the representiveness of the data, this paper use the multiple-imputation and dropping methods [@gelman]. There are two types of missing values: numercial and categorical missing values.

@@

The number of oberservation before data cleaning is 4032. This number decrease to 3794 because some rows with missing values. are dropped from the data.

In addtion, the variable of **degree** is grouped into two levels. For respondents who have obtained at least a degree from an associate/junior college, their levels are denoted as "1", indicating they have complete a post-secondary education. For respondents who have attained only a high school diploma or possess a lower level of education, their levels are designated as "0".

@@

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-degree
#| tbl-cap: Number of respondents by degree for 2021 survey
tb1 <- cleaned_data %>% count(degree) %>%
  mutate(percent=percent(x=n/sum(n)))
tb1<-tb1%>%arrange(factor(tb1$degree,c("graduate", "bachelor's", "associate/junior college", "high school", "less than high school")))
colnames(tb1) <- c("Degree", "Total", "Proportion")
tb1%>%kable(booktabs = TRUE)
```
According to the @tbl-degree, the distribution of education levels in terms of degrees shows that the most prevalent category is high school education, with 1507 individuals (39.7%) falling into this group. Next in prevalence is a bachelor's degree, held by 1010 individuals (26.6%). A graduate degree is the third most common category, with 731 individuals (19.3%) having attained this level of education. In contrast, the associate/junior college degree is less frequent, with 362 individuals (9.5%) completing this level of education. The smallest proportion, 4.8%, have less than a high school education.

```{r}
#| label: fig-educ
#| fig-cap: Histogram of Respondents' Education
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 4
ggplot(data = cleaned_data, aes(x = educ)) +

@@

  # geom_histogram(bins = 15, fill = "lightblue", color = "white") +
  labs(x = "Education (in Years)", y = "Density") +
  theme_bw()
```
The second variable utilized to quantify education is relation to the number of years of schooling. In @fig-educ, this histogram shows an overall distribution of the respondents' education in years. It is clear that the distribution is right-skewed, which indicates that there are more respondents with higher levels of education. In other words, the bulk of the respondents are clustered towards the left side of the histogram (lower levels of education), and there are relatively fewer respondents towards the right side of the histogram (higher levels of education). In short, the majority of respondents in 2021 GSS are highly educated.

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-educ
#| tbl-cap: Number of respondents by education for 2021 survey
tb0 <- cleaned_data %>% summarise(Min = min(cleaned_data$educ,na.rm=T),
                              Median = median(cleaned_data$educ,na.rm=T),
                              Max = max(cleaned_data$educ,na.rm = T),
                              Mean = round(mean(cleaned_data$educ,na.rm=T),digits=2))

tb0 %>% kable()
```
To see a more accurate numerical summary, @tbl-educ is created. In the dataset labelled @sec-data, the category denoting "no formal schooling" has been redefined as an integer value, 0. According to the information presented in the @tbl-educ table, the range of years of schooling reported by respondents spans from 0 to 20 years. The median number of years of schooling reported is 15, and the mean is 14.82 This mean value roughly corresponds to three years in college. Taken together, these findings suggest that the respondents surveyed in the 2021 GSS have generally completed their high school education and are highly educated.

```{r}
#| label: fig-parentpre
#| fig-cap: Mother's and Father's Prestige Score
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 4
# ggplot(cleaned_data, aes(x = mapres10)) +
#   geom_histogram(aes(y = ..density.., fill = "mapres10"), alpha = 1, binwidth = 2,color = "black") +
#   geom_histogram(aes(x = papres10, y = ..density.., fill = "paeduc"), alpha = 0.8, binwidth = 2,color = "black") +
#   scale_fill_manual(values = c("mapres10" = "lightblue", "paeduc" = "darkred"), name = "Variables") +
#   labs(title = "Mother's and Father's Prestige Score",
#        x = "Prestige Score",
#        y = "Density")+
#   theme_bw()

#prestige score distribution
ggplot(cleaned_data, aes(x = mapres10)) +
  geom_histogram(aes(y = ..density.., fill = "Mother's Prestige Score"), alpha = 1, binwidth = 1, color = "black")+
  geom_histogram(aes(x = papres10, y = ..density.., fill = "Father's Prestige Score"), alpha = 0.8, binwidth = 1, color = "black") +
  scale_fill_manual(values = c("Mother's Prestige Score" = "darkgreen", "Father's Prestige Score" = "lightgreen"), name = "Variables") +
  labs(x = "Prestige Score",
       y = "Density") +
  theme_bw()
```
In @fig-parentpre, it shows the distribution of respondent’s father’s and mother’s prestige scores. The distribution of father’s prestige score and the mother’s prestige score have quite different distributions. Specifically, the distribution of mother's prestige score is bimodal and the distribution of father's prestige score is unimodal.

```{r}
#| label: fig-presandeduc
#| fig-cap: Mother's and Father's Prestige Score with children's education
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 4
#pres vs. educ
p1 <- ggplot(cleaned_data, aes(x=educ, y=papres10)) +
  geom_point(color="#69b3a2", size=2,alpha=0.3) +
  labs(x="Education (in Years)", y="Father's Prestige Score")+
  theme_bw()
p2 <- ggplot(cleaned_data, aes(x=educ, y=mapres10)) +
  geom_point(color="darkgreen",size=2,alpha=0.3) +
  labs(x="Education (in Years)", y="Mother's Prestige Score")+
  theme_bw()

p1 + p2
```
In order to examine the relationship between parents’ prestige scores and children’s education level. The next plot, @fig-presandeduc, is used to show the relation. These two scatter plots show the father’s prestige score and the mother’s prestige score with their children’s education. It is hard to see a clear relation or pattern based on these two scatter plots. To verify if the prestige score is one of the factors, a linear model will be used in the @sec-model.

```{r}
#| label: fig-parentseduc
#| fig-cap: Mother's Education and Father's Education Level
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 4
ggplot(cleaned_data, aes(x = maeduc)) +
  geom_histogram(aes(y = ..density.., fill = "Mother's Education"),
                 alpha = 1, binwidth = 0.5, color ="black") +geom_histogram(aes(x = paeduc, y = ..density.., fill = "Father's Education"),alpha = 0.8, binwidth = 0.5, color = "black") +
scale_fill_manual(values = c("Mother's Education" = "darkgreen", "Father's Education" = "lightgreen"), name = "Variables") +
labs(x = "Education (in Years)",
y = "Density") +
theme_bw()
```
Another key variable is the parent’s education level. In @fig-parentseduc, it shows the distribution of the mother’s education and the father’s education. One interesting result is that the distribution of education for mother and father is similar, or even almost identical. This indicates that there is a strong correlation between the education levels of mothers and fathers. In other words, if a mother has a high level of education, it is likely that the father also has a high level of education, and vice versa.

@@

ggplot(cleaned_data, aes(x = educ, fill = factor(born))) +
  geom_histogram(aes(y = ..density..), position = "dodge", binwidth = 1.5, alpha = 0.5, color = "black")+
   labs(x = "Education (in Years)",
       y = "Density",
       fill = "Birthplace of the Respondent (if in US)")+
  theme_bw()

# ggplot(cleaned_data, aes(x = educ)) +
# geom_histogram(aes(y = ..density.., fill = "Mother's Education"), alpha = 1, binwidth = 0.5, color = "black") +
# geom_histogram(aes(x = paeduc, y = ..density.., fill = "Father's Education"), alpha = 0.8, binwidth = 0.5, color = "black") +
# scale_fill_manual(values = c("Mother's Education" = "darkgreen", "Father's Education" = "lightgreen"), name = "Variables") +
# labs(x = "Education (in Years)",
# y = "Density") +
# theme_bw()
```
Based on [@overman], the birthplace may have an impact on children’s education. In @fig-birth, it is used to test the relationship between birthplace and children’s education. Note that the question of birthplace only asks if the respondent was born in the US or not. In general, most of the respondents have received at least 10 years of education despite their birthplace. However, the bar for respondents who were born outside of the US is much higher than the other groups. This is because 90% of the respondents were born in the US, and only 10.0% answered “No”. This is because respondents who did not answer this question or answered NA are removed during the data cleaning. Given the huge differences in the sample sizes between groups, it is hard to draw a general and representative conclusion of whether birthplace may affect the education level.

```{r,echo=FALSE}
tb3 <- cleaned_data %>% 
  count(born) %>% 
  mutate(percent=percent(x=n/sum(n)))
```

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-regandeduc
#| tbl-cap: The proportion and mean of education for different regions
tb_reg <- cleaned_data %>% 
  count(reg16) %>% 
  mutate(percent=percent(x=n/sum(n)))

tb_reg_avg<-cleaned_data %>%
  group_by(reg16) %>%
  summarise(mean_educ = round(mean(educ),digits=2))
tb_reg<-tb_reg%>%mutate("Mean of Education"=tb_reg_avg$mean_educ)
colnames(tb_reg)<-c("Region","Count","Percent","Mean of Education")
tb_reg %>% kable()
```
The @tb-regandeduc showcases the average number of years of education for individuals in different living regions of the United States when the respondent's was 16. The highest average education level is found in the Middle Atlantic region, where the mean is 15.13 years. The East North Central region follows closely, with an average of 14.98 years of education. In the Pacific and New England regions, the mean years of education are relatively similar, with 14.94 and 14.85 years, respectively. Those regions are above the overall average, which is 14.82 from @tb_educ. On the other hand, the South Atlantic and West North Central regions also display comparable average education levels of 14.75 and 14.73 years, respectively. The Mountain region exhibits a slightly lower mean of 14.69 years. Furthermore, the East South Atlantic and West South Central regions have the lowest average education levels, with 14.10 and 14.39 years, respectively.

```{r}
#| label: fig-region
#| fig-cap: Respondent's Living Region at Age of 16 v.s. Education
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 4
ggplot(cleaned_data, aes(x = educ,fill = factor(reg16))) +
  geom_histogram(aes(y = ..density..),binwidth = 2, color = "black") +
  facet_wrap(~ reg16, ncol = 3)+
  labs(x = "Education (in Years)",
       y = "Density")+
  theme_bw()+
  theme(legend.position = "none")
```
According to the study [@nieuwenhuis], there appears to be an association between neighborhood and education. In the 2021 GSS, the neighbourhood variable is being measured by the living region at the age of 16. The distribution for all regions, as shown in @fig-region, follows a similar bimodal shape. This could indicate that the region of living has a minimal or no impact on children's education levels. However, it is important to note that the measurement of the neighborhood through living regions at age 16 may not fully capture the impact of the neighborhood on education, which will be discussed in the @sec-weakness.

```{r}
#| label: fig-maincome
#| fig-cap: Education vs Mother's Education with respect to Family Income
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 4
ggplot(cleaned_data, aes(x = educ, y = maeduc, color = incom16)) +
  geom_jitter(alpha = 0.2) +
  stat_smooth(method = "lm", se = FALSE) +
  facet_wrap(~incom16)+
  labs(x = "Education (in Years)",
         y = "Mother's Education (in Years)",
       color = "Family Income (at Age of 16)") +
  scale_color_discrete(name = "Family Income (at Age of 16)") +
  theme_bw()
```

To examine the relationship between family income and education, the @fig-maincome and @fig-paincome visualize this relation. In @fig-maincome, the vertical axis represents the mother’s education and the horizontal axis represents the children’s education. The respondents were asked to rate their family income at age of 16 in the following category: (1) **far above average**, (2) **above average**, (3) **average**, (4) **below average** and (5) **far below average**. Overall, mothers’ education and children’s education are positively correlated despite the level of income. To be more specific, for respondents whose family income was **far above average**, the slope seems to be steeper than the other groups, which indicates that for those with higher family income, there may be a stronger relationship between the education of the mother and the education of the children.

@@

ggplot(cleaned_data, aes(x = educ, y = paeduc, color = incom16)) +
  geom_jitter(alpha = 0.2) +
  stat_smooth(method = "lm", se = FALSE) +
  facet_wrap(~incom16)+
  labs(x = "Education (in Years)",
       y = "Father's Education (in Years)",
  color = "Family Income (at Age of 16)") +
  scale_color_discrete(name = "Family Income (at Age of 16)") +
  theme_bw()
```
In @fig-paincome, the income groups are the same as before. However, the plot is now examining the correlation between the father’s education and the children’s education. Once again, there is a positive correlation between the education level of fathers and their children's education, regardless of family income. However, for respondents who reported a **far above average** family income, the relationship between the father's education and children's education appears to be stronger than for other income groups. This suggests that higher family income may be associated with a more pronounced link between paternal and children's education levels. Comparing the two plots, the slopes in @fig-paincome for all income groups are much steeper than the slopes in @fig-maincome. This may suggest that there is a stronger relationship between fathers' education and children's education compared to mothers' education and children's education given the same income level.

# Model* {#sec-model}

$$
\log(\frac{\pi}{1-\pi})=\hat{\beta_0}+\hat{\beta_1}X_{mapres10}+\hat{\beta_2}X_{papres10}+\hat{\beta_3}X_{paeduc}+\hat{\beta_4}X_{maeduc}
$$ {#eq-model}

>>>>

---
title: "Exploring the Relationship between Family Background and Children's Education Level in the United States (2021)"
author: 
  - Ruibo Sun
thanks: "Code and Data: https://github.com/ruibosun/revised-how-parents-affect-childrens-education"
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
abstract: "This paper explores the relationship between family status variables, such as parents' socio-economic status and educational level, and their impact on children's academic achievement and education level using the data from 2021 General Social Survey (GSS). Previous studies have suggested that parents' education can positively influence their children's academic performance due to increased access to resources and greater involvement in their education. However, recent research has demonstrated that this relationship is complex and can be influenced by a variety of factors. The findings in this paper suggest that parents' education, family income, and prestige score all affect the children's education level. Further research is needed to fully understand the complexities of this relationship and use the results to improve educational policies and programs."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
library(broom)
library(knitr)
library(kableExtra)
library(tidyverse)
library(dplyr)
library(here)
library(haven)
library(scales)
library(patchwork)
```

@@

In short, the findings in this paper demonstrate that parents' education has a positive correlation with children's education. Additionally, the region of living during the teenage years can also impact children's education. Parents' prestige score seems to be another factor in children's education. However, there is some limitation on the definition of prestige score that may influence the overall results. This will be discussed in the @sec-weakness.

The data will be presented clearly and succinctly with plots and tables. This following analysis is processed in `R` [@R] with packages `tidyverse` [@tidyverse], `dplyr` [@dplyr], `here` [@here], `haven` [@haven] and `broom` [@broom]. The tables are constructed via `knitr` [@knitr], `scales` [@scales] and kableExtra [@kableExtra]. The package inside `tidyverse` helps to create the plots in `ggplot2` [@tidyverse]. The missing values were imputed using `mice` [@mice]. This paper is knitted as a PDF file by the packages of Quarto and `patchwork` [@patchwork].

@@

The target population is US adults aged 18 and older. The GSS uses a multistage probability sampling approach to recruit its sample from the US Census Bureau's Master Address File. One strength of the GSS is its long history and large sample size, but a potential weakness is a reliance on self-reported data, which can be subject to response bias. The questionnaire includes both closed-ended and open-ended questions, but there may be concerns about biased or leading questions. The GSS employs various methods to adjust for non-response and ensure the representativeness of the sample. The questionnaire includes both closed-ended and open-ended questions, but there may be concerns about biased or leading questions. [@norcuofch]. The GSS employs various methods to adjust for non-response and ensure the representativeness of the sample. The limitations will be explained in detail in @sec-weakness.

$$
\mbox{Prestige}_{ij}=\mu+\alpha_i+\beta_{j}+\epsilon_{ij}
$$ {#eq-modpr}

This paper will test and explore how parents' social-economics status can affect children's education. The data that will be used in this paper comes from the US General Social Survey from the National Opinion Research Center at the University of Chicago. In this study, the "children" are the respondents and "parent" are the parents of the respondents.

The factor of socio-economics status is measured using the occupation prestige score. It is a measure used in social science research to assess the level of social status or prestige associated with a particular occupation or profession, which was developed by the National Opinion Research Center at the University of Chicago. As part of the GSS, respondents are asked to rate the prestige or social standing of various occupations on a scale of 1 to 100, with higher scores indicating greater prestige. The ideal dataset would have 90,090 ratings from 1,001 raters, but due to various issues, only 82,800 ratings from 946 raters were used. Invalid or inconsistent ratings, such as cases with reversed codes, too few ratings, equal scores for all occupations, or a low standard deviation of ratings, were excluded to ensure the reliability of the occupational scores [@smith].

Next, the prestige score is obtained by the @eq-modpr. $\mbox{Prestige}_{ij}$ is the prestige score for occupational title $i$ and rater $j$. The $\alpha_i$ represents the occupational differences of interest, while the $\beta_j$ represents differences among the raters that need to be controlled for in estimating the $\alpha_i$. The $\mu$ is the overall mean prestige score, and $\epsilon_{ij}$ represents the random error term. The range of the prestige scores is from 0 to 100, with higher scores indicating a better or more prestigious occupation [@smith].

@@

- **mapres10**: A numeric variable indicating the respondent's mother's prestige score.
- **papres10**: A numeric variable indicating the respondent's father's prestige score.
- **paeduc**: A numeric variable indicating the number of years of education completed by the respondent's father.
- **maeduc**: A numeric variable indicating the number of years of education completed by the respondent's mother.
- **educ**: A numeric variable indicating the number of years of education completed by the respondent.
- **degree**: A categorical variable indicating the highest degree earned by the respondent.
- **born**: A categorical variable indicating the region of the US where the respondent was born.
- **reg16**: A categorical variable indicating the region of the US where the respondent lived at age of 16.
- **incom16**: A categorical variable indicating the respondent's total household at age of 16.

@@

As mentioned before, this paper is an revised version of [the previous paper](https://github.com/ruibosun/how-parents-affect-childrens-education/blob/main/outputs/paper/paper.pdf) which has some limitations and weaknesses. In order to improve the representiveness of the data, this paper use the multiple-imputation and dropping methods [@gelman]. There are two types of missing values: numerical and categorical missing values.

@@

The number of observations before data cleaning is 4032. This number decrease to 3794 because some rows with missing values. are dropped from the data.

In addition, the variable of **degree** is grouped into two levels. For respondents who have obtained at least a degree from an associate/junior college, their levels are denoted as "1", indicating they have completed a post-secondary education. For respondents who have attained only a high school diploma or possess a lower level of education, their levels are designated as "0".

@@

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-degree
#| tbl-cap: Number of respondents by degree for 2021 survey

tb1 <- cleaned_data %>% count(degree) %>%
  mutate(percent = percent(x = n / sum(n)))

tb1 <-
  tb1 %>% arrange(factor(
    tb1$degree,
    c(
      "graduate",
      "bachelor's",
      "associate/junior college",
      "high school",
      "less than high school"
    )
  ))

colnames(tb1) <- c("Degree", "Total", "Proportion")

tb1 %>% kable(booktabs = TRUE)
```
According to the @tbl-degree, the distribution of education levels in terms of degrees shows that the most prevalent category is high school education, with 1507 individuals (39.7%) falling into this group. Next in prevalence is a bachelor's degree, held by 1010 individuals (26.6%). A graduate degree is the third most common category, with 731 individuals (19.3%) having attained this level of education. In contrast, the associate/junior college degree is less frequent, with 362 individuals (9.5%) completing this level of education. The smallest proportion, 4.8%, have less than a high school education.

```{r}
#| label: fig-educ
#| fig-cap: Histogram of Respondents' Education
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 4

ggplot(data = cleaned_data, aes(x = educ)) +

@@

  # geom_histogram(bins = 15, fill = "lightblue", color = "white") +
  labs(x = "Education (in Years)", y = "Density") +
  theme_bw()
```

The second variable utilized to quantify education is relation to the number of years of schooling. @fig-educ shows an overall distribution of the respondents' education in years. It is clear that the distribution is right-skewed, which indicates that there are more respondents with higher levels of education. In other words, the bulk of the respondents are clustered towards the left side of the histogram (lower levels of education), and there are relatively fewer respondents towards the right side of the histogram (higher levels of education). In short, the majority of respondents in 2021 GSS are highly educated.

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-educ
#| tbl-cap: Number of respondents by education for 2021 survey

tb0 <-
  cleaned_data %>% summarise(
    Min = min(cleaned_data$educ, na.rm = T),
    Median = median(cleaned_data$educ, na.rm =
                      T),
    Max = max(cleaned_data$educ, na.rm = T),
    Mean = round(mean(cleaned_data$educ, na.rm =
                        T), digits = 2)
  )

tb0 %>% kable()
```

To see a more accurate numerical summary, @tbl-educ is created. In the dataset labelled @sec-data, the category denoting "no formal schooling" has been redefined as an integer value, 0. According to the information presented in the @tbl-educ table, the range of years of schooling reported by respondents spans from 0 to 20 years. The median number of years of schooling reported is 15, and the mean is 14.82 This mean value roughly corresponds to three years in college. Taken together, these findings suggest that the respondents surveyed in the 2021 GSS have generally completed their high school education and are highly educated.

```{r}
#| label: fig-parentpre
#| fig-cap: Mother's and Father's Prestige Score
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 4

#prestige score distribution
ggplot(cleaned_data, aes(x = mapres10)) +
  geom_histogram(
    aes(y = ..density.., fill = "Mother's Prestige Score"),
    alpha = 1,
    binwidth = 1,
    color = "black"
  ) +
  geom_histogram(
    aes(x = papres10, y = ..density.., fill = "Father's Prestige Score"),
    alpha = 0.8,
    binwidth = 1,
    color = "black"
  ) +
  scale_fill_manual(
    values = c(
      "Mother's Prestige Score" = "darkgreen",
      "Father's Prestige Score" = "lightgreen"
    ),
    name = "Variables"
  ) +
  labs(x = "Prestige Score",
       y = "Density") +
  theme_bw()
```

@fig-parentpre shows the distribution of respondent's father's and mother's prestige scores. The distribution of father's prestige score and the mother's prestige score have quite different distributions. Specifically, the distribution of mother's prestige score is bimodal and the distribution of father's prestige score is unimodal.

```{r}
#| label: fig-presandeduc
#| fig-cap: Mother's and Father's Prestige Score with children's education
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 4

#pres vs. educ
p1 <- ggplot(cleaned_data, aes(x = educ, y = papres10)) +
  geom_point(color = "#69b3a2",
             size = 2,
             alpha = 0.3) +
  labs(x = "Education (in Years)", y = "Father's Prestige Score") +
  theme_bw()
p2 <- ggplot(cleaned_data, aes(x = educ, y = mapres10)) +
  geom_point(color = "darkgreen",
             size = 2,
             alpha = 0.3) +
  labs(x = "Education (in Years)", y = "Mother's Prestige Score") +
  theme_bw()

p1 + p2
```
In order to examine the relationship between parents' prestige scores and children's education level. The next plot, @fig-presandeduc, is used to show the relation. These two scatter plots show the father's prestige score and the mother's prestige score with their children's education. It is hard to see a clear relation or pattern based on these two scatter plots. To verify if the prestige score is one of the factors, a linear model will be used in the @sec-model.

```{r}
#| label: fig-parentseduc
#| fig-cap: Mother's Education and Father's Education Level
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 4

ggplot(cleaned_data, aes(x = maeduc)) +
  geom_histogram(
    aes(y = ..density.., fill = "Mother's Education"),
    alpha = 1,
    binwidth = 0.5,
    color = "black"
  ) + geom_histogram(
    aes(x = paeduc, y = ..density.., fill = "Father's Education"),
    alpha = 0.8,
    binwidth = 0.5,
    color = "black"
  ) +
  scale_fill_manual(
    values = c(
      "Mother's Education" = "darkgreen",
      "Father's Education" = "lightgreen"
    ),
    name = "Variables"
  ) +
  labs(x = "Education (in Years)",
       y = "Density") +
  theme_bw()
```

Another key variable is the parent's education level. In @fig-parentseduc, it shows the distribution of the mother's education and the father's education. One interesting result is that the distribution of education for mother and father is similar, or even almost identical. This indicates that there is a strong correlation between the education levels of mothers and fathers. In other words, if a mother has a high level of education, it is likely that the father also has a high level of education, and vice versa.

@@

ggplot(cleaned_data, aes(x = educ, fill = factor(born))) +
  geom_histogram(
    aes(y = ..density..),
    position = "dodge",
    binwidth = 1.5,
    alpha = 0.5,
    color = "black"
  ) +
  labs(x = "Education (in Years)",
       y = "Density",
       fill = "Birthplace of the Respondent (if in US)") +
  theme_bw()
```

@overman suggests birthplace may have an impact on children's education. In @fig-birth, it is used to test the relationship between birthplace and children's education. Note that the question of birthplace only asks if the respondent was born in the US or not. In general, most of the respondents have received at least 10 years of education despite their birthplace. However, the bar for respondents who were born outside of the US is much higher than the other groups. This is because 90% of the respondents were born in the US, and only 10.0% answered “No”. This is because respondents who did not answer this question or answered NA are removed during the data cleaning. Given the huge differences in the sample sizes between groups, it is hard to draw a general and representative conclusion of whether birthplace may affect the education level.

```{r}
#| echo: false
tb3 <- cleaned_data %>%
  count(born) %>%
  mutate(percent = percent(x = n / sum(n)))
```

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-regandeduc
#| tbl-cap: The proportion and mean of education for different regions

tb_reg <- cleaned_data %>%
  count(reg16) %>%
  mutate(percent = percent(x = n / sum(n)))

tb_reg_avg <- cleaned_data %>%
  group_by(reg16) %>%
  summarise(mean_educ = round(mean(educ), digits = 2))
tb_reg <- tb_reg %>% mutate("Mean of Education" = tb_reg_avg$mean_educ)
colnames(tb_reg) <- c("Region", "Count", "Percent", "Mean of Education")
tb_reg %>% kable()
```

@tb-regandeduc showcases the average number of years of education for individuals in different living regions of the United States when the respondent was 16. The highest average education level is found in the Middle Atlantic region, where the mean is 15.13 years. The East North Central region follows closely, with an average of 14.98 years of education. In the Pacific and New England regions, the mean years of education are relatively similar, with 14.94 and 14.85 years, respectively. Those regions are above the overall average, which is 14.82 from @tb_educ. On the other hand, the South Atlantic and West North Central regions also display comparable average education levels of 14.75 and 14.73 years, respectively. The Mountain region exhibits a slightly lower mean of 14.69 years. Furthermore, the East South Atlantic and West South Central regions have the lowest average education levels, with 14.10 and 14.39 years, respectively.

```{r}
#| label: fig-region
#| fig-cap: Respondent's Living Region at Age of 16 v.s. Education
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 4

ggplot(cleaned_data, aes(x = educ, fill = factor(reg16))) +
  geom_histogram(aes(y = ..density..),
                 binwidth = 2,
                 color = "black") +
  facet_wrap( ~ reg16, ncol = 3) +
  labs(x = "Education (in Years)",
       y = "Density") +
  theme_bw() +
  theme(legend.position = "none")
```

@nieuwenhuis finds an association between neighborhood and education. In the 2021 GSS, the neighbourhood variable is being measured by the living region at the age of 16. The distribution for all regions, as shown in @fig-region, follows a similar bimodal shape. This could indicate that the region of living has a minimal or no impact on children's education levels. However, it is important to note that the measurement of the neighborhood through living regions at age 16 may not fully capture the impact of the neighborhood on education, which will be discussed in the @sec-weakness.

```{r}
#| label: fig-maincome
#| fig-cap: Education vs Mother's Education with respect to Family Income
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 4

ggplot(cleaned_data, aes(x = educ, y = maeduc, color = incom16)) +
  geom_jitter(alpha = 0.2) +
  stat_smooth(method = "lm", se = FALSE) +
  facet_wrap( ~ incom16) +
  labs(x = "Education (in Years)",
       y = "Mother's Education (in Years)",
       color = "Family Income (at Age of 16)") +
  scale_color_discrete(name = "Family Income (at Age of 16)") +
  theme_bw()
```

To examine the relationship between family income and education, the @fig-maincome and @fig-paincome visualize this relation. In @fig-maincome, the vertical axis represents the mother's education and the horizontal axis represents the children's education. The respondents were asked to rate their family income at age of 16 in the following category: (1) **far above average**, (2) **above average**, (3) **average**, (4) **below average** and (5) **far below average**. Overall, mothers' education and children's education are positively correlated despite the level of income. To be more specific, for respondents whose family income was **far above average**, the slope seems to be steeper than the other groups, which indicates that for those with higher family income, there may be a stronger relationship between the education of the mother and the education of the children.

@@

ggplot(cleaned_data, aes(x = educ, y = paeduc, color = incom16)) +
  geom_jitter(alpha = 0.2) +
  stat_smooth(method = "lm", se = FALSE) +
  facet_wrap( ~ incom16) +
  labs(x = "Education (in Years)",
       y = "Father's Education (in Years)",
       color = "Family Income (at Age of 16)") +
  scale_color_discrete(name = "Family Income (at Age of 16)") +
  theme_bw()
```

In @fig-paincome, the income groups are the same as before. However, the plot is now examining the correlation between the father's education and the children's education. Once again, there is a positive correlation between the education level of fathers and their children's education, regardless of family income. However, for respondents who reported a **far above average** family income, the relationship between the father's education and children's education appears to be stronger than for other income groups. This suggests that higher family income may be associated with a more pronounced link between paternal and children's education levels. Comparing the two plots, the slopes in @fig-paincome for all income groups are much steeper than the slopes in @fig-maincome. This may suggest that there is a stronger relationship between fathers' education and children's education compared to mothers' education and children's education given the same income level.

# Model {#sec-model}

We estimate the following mode:
